Probabilistic Programming around Gaussian Processes

### 2019-10-21 abstract/paper.tex
drafted, typo fixed

\begin{abstract}
    We present GoGP, a library for probabilistic programming
    around Gaussian processes. Kernels, beliefs about
    hyperparameters and about observation inputs and outputs are
    programmatically expressed using the same differentiable
    probabilistic programming framework. On one hand, a basic
    usage with maximum-likelihood hyperparameter estimates and
    homoscedastic Gaussian noise requires as little coding as
    with any Gaussian process library. On the other hand,
    imposing prior beliefs on hyperparameters, handling input
    uncertainty, heteroscedastic and non-Gaussian noise, or
    change point detection can be organically added on top of a
    Gaussian process kernel. Just like distributions in
    probabilistic programs, kernels can be flexibly combined
    using general code flow rather than through predefined
    combination operators (usually addition and multiplication).
    Implemented in Go, a general-purpose programming language
    popular for server-side programming, our library is both
    well suited for integration into data processing/machine
    learning pipelines and exploits language features such as
    efficient light-weight computation parallelism for
    computation efficiency.  We demonstrate different uses of
    the library on a variety of case studies inspired by
    applications. 
\end{abstract}

@@

Gaussian processes are successfully deployed in many settings,
in particular when 
\begin{itemize}
    \item a stationary kernel is sufficient;
    \item kernel hyperparameters can be fixed \textit{a priori} or
    	reliably estimated from observations;
    \item observation noise is constant.
\end{itemize}
Most basic cases are indeed covered by the above assumptions,
and library implementations of Gaussian processes prioritize
scalability with number of observations, rich libraries
of kernel functions, and integration of Gaussian processes with
machine learning pipelines.

There are cases though where the noise varies between
observations, structured informative priors must be imposed on
kernel hyperparameters, observation inputs are uncertain, or
observations are best explained by presence of change
points~\cite{}.

@@

GoGP is built around a dual view on the Gaussian process: as a
stochastic process and as a probabilistic model with respect
to kernel hyperparameters and, optionally, observations.
\lstinline{GP}, the type encapsulating a Gaussian process,
provides methods \lstinline{Absorb} and \lstinline{Produce} to 
add observations to the process and predicting the distribution
of outputs at given locations. In addition, \lstinline{GP}
provides methods \lstinline{Observe} and \lstinline{Gradient}
that turn it into an Infergo model (with a custom gradient).
Inference can be performed on a \lstinline{GP} instance alone
or on a model that combines a \lstinline{GP} instance with
beliefs about hyperparameters and observations.  The
log-likelihood gradient is computed with respect to
hyperparameters only or with respect to both hyperparameters and
observations, supporting both basic and advanced usage.

@@

The library user specifies the kernel:
\begin{lstlisting}
type Basic struct{}
func (Basic) Observe(x []float64) float64 {
    return x[0] * kernel.Normal.Observe(x[1:])
}
func (Basic) NTheta() int { return 2 }
\end{lstlisting}
and initializes \lstinline{GP} with a kernel instance:
\begin{lstlisting}
gp := &gp.GP{
    NDim:  1,
    Simil: &Basic{},
}
\end{lstlisting}

@@

In basic usage, the observation noise is assumed to be Gaussian.
This usage is supported by initializing \lstinline{GP} with a noise
kernel, along with a similarity kernel. When the noise is not
Gaussian, an analytical solution for posterior Gaussian process
inference does not always exist. However, non-Gaussian noise is
straightforwardly supported by GoGP through supplying a
model for beliefs on observation outputs:
\begin{lstlisting}
type Noise struct {
	Y []float64 // noisy outputs
}
func (m *Noise) Observe(x []float64) float64 {
	// Laplacian noise
	for i := range m.Y {
		ll += Expon.Logp(1/math.Exp(x[s]),
			math.Abs(m.Y[i]-x[i0+n+i]))
	}
	return ll
}
\end{lstlisting}

>>>>

\begin{abstract}
    We present GoGP, a library for probabilistic programming
    around Gaussian processes. Kernels, beliefs about
    hyperparameters and about observation inputs and outputs are
    programmatically expressed using the same differentiable
    probabilistic programming framework. On one hand, a basic
    usage with maximum-likelihood hyperparameter estimates and
    homoscedastic Gaussian noise requires as little coding as
    with any Gaussian process library. On the other hand,
    imposing prior beliefs on hyperparameters, handling input
    uncertainty, heteroscedastic and non-Gaussian noise, or
    change point detection can be organically added on top of a
    Gaussian process kernel. Just like distributions in
    probabilistic programs, kernels can be flexibly combined
    using general code flow rather than through predefined
    combination operators (usually addition and multiplication).
    Implemented in Go, a general-purpose programming language
    popular for server-side programming, our library is both
    well suited for integration into data processing/machine
    learning pipelines and exploits language features such as
	efficient light-weight parallelism for computation
	efficiency.  We demonstrate different uses of the library on
	a variety of case studies inspired by applications. 
\end{abstract}

@@

Gaussian processes are successfully deployed in many settings,
in particular when 
\begin{itemize}
    \item a stationary kernel is sufficient;
    \item kernel hyperparameters can be fixed \textit{a priori} or
    	reliably estimated from observations;
    \item observation noise is constant.
\end{itemize}
Most basic cases are indeed covered by the above assumptions,
and library implementations of Gaussian processes prioritize
scalability with number of observations, rich libraries
of combinable kernel functions, and integration of Gaussian
processes with machine learning pipelines.

There are cases though where noise varies between
observations, structured informative priors must be imposed on
kernel hyperparameters, observation inputs are uncertain, or
observations are best explained by presence of change
points~\cite{}.

@@

GoGP is built around a dual view on the Gaussian process: as a
stochastic process and as a probabilistic model with respect
to kernel hyperparameters and, optionally, observations.
\lstinline{GP}, the type encapsulating a Gaussian process,
provides methods \lstinline{Absorb} and \lstinline{Produce} to 
add observations to the process and to predict the distribution
of outputs at given locations. In addition, \lstinline{GP}
provides methods \lstinline{Observe} and \lstinline{Gradient}
that turn it into an Infergo model (with a custom gradient).
Inference can be performed on a \lstinline{GP} instance alone
or on a model that combines a \lstinline{GP} instance with
beliefs about hyperparameters and observations.  The
log-likelihood gradient is computed with respect to
hyperparameters only or with respect to both hyperparameters and
observations, supporting both basic and advanced usage.

@@

The library user specifies the kernel:
\begin{lstlisting}
type Basic struct{}
func (Basic) Observe(x []float64) float64 {
    return x[0] * kernel.Normal.Observe(x[1:])
}
func (Basic) NTheta() int { return 2 }
\end{lstlisting}
and initializes \lstinline{GP} with a kernel instance:
\begin{lstlisting}
gp := &gp.GP{
    NDim:  1,
    Simil: Basic{},
}
\end{lstlisting}

@@

In basic usage, the observation noise is assumed to be Gaussian.
This usage is supported by initializing \lstinline{GP} with a noise
kernel, along with a similarity kernel. When the noise is not
Gaussian, an analytical solution for posterior Gaussian process
inference does not always exist. However, non-Gaussian noise is
straightforwardly supported by GoGP through supplying a
model for beliefs on observation outputs:
\begin{lstlisting}
type Noise struct {
	Y []float64 // noisy outputs
}
func (m *Noise) Observe(x []float64) float64 {
	// Laplacian noise
	for i := range m.Y {
		ll += Expon.Logp(1/math.Exp(x[0]),
			math.Abs(m.Y[i]-x[1+i]))
	}
	return ll
}
\end{lstlisting}
